---
title: "Penumonia SAR Models"
output:
  html_document:
    keep_md: yes
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    number_sections: true
---

This is a reproducible example of the paper [Pneumonia Sar](). In this research, we perform a spatial autorregresive model analysis to measure the spatial influence of socio-economic, enviromental, behavioural and healthcare factors in pneumonia mortality rates. 

# Getting Started

## Packages

First, we will load the packages needed for processing and analyzing our data.

```{r}
# Omitting warmings (be carefull)
options(warn = -1)

# Packages
packages <- c('lmtest','RColorBrewer','classInt','spdep','TeachingDemos','shapefiles','sp','maptools',
             'scatterplot3d','geoR','spatial','fBasics','car','aplpack','RODBC','ggplot2','spgrass6',
             'adespatial','RANN','ade4','olsrr','rgeos','rgdal','spdep','spgwr','GWmodel','nnet','olsrr',
             'stats','classInt','gridExtra','lmtest','car','MASS','caret','glmnet')
```

For loading and/or installing the packages, you can run the following command. The packages that are installed in your machine will be automatically loaded. Those that are not installed will be installed and loaded to your R session. Remember to provide permits if linux is used.

```{r, results=FALSE}
# Permits
#Sys.setenv(R_INSTALL_STAGED = FALSE)

# Load packages
for(p in packages){
  if(!require(p,character.only = TRUE)) install.packages(p, dependencies = TRUE)
  library(p,character.only = TRUE)
}
```

## Working directory

It is easier to work in R using a working directory. That way you ensure your inputs and outputs will be loaded and stored in a global directory.

```{r}
# Set working directory
dir <- '/mnt/d/M.Sc. Gesopatial Tecnologies/GeoMundus/GeoMundus 2019/Neumonia'
setwd(file.path(dir))
```

We will also use some useful functions in the exploratory analisis. Load them all from the *SpatialFunctions* folder. Remember this folder has to be in your current working directory.

```{r}
# Load some useful functions
funDir <- file.path(dir, 'SpatialFunctions')
spFun <- list.files(funDir, pattern = '.r', full.names = T)
invisible(lapply(spFun, source))
```

And we will add an extra function for resting the plot options everytime we plot.

```{r}
# Reset par for sequential plotting
resetPar <- function() { dev.new(); op <- par(no.readonly = TRUE); dev.off(); op}
```

# Data

Our data contains the [standardized mortality ratio (SMR)](https://ibis.health.state.nm.us/resource/SMR_ISR.html#:~:text=Standardized%20Mortality%20Ratio%20(SMR)%20is,the%20same%20age%2Fsex%20groups.) of Pneumonia (our study variable) in 19 districts of Bogotá, Colombia for the years 2004, 2007, 2011 and 2014. The data also contains socio-economic, enviromental and healthcare covariates.

As our data is stored in a *shapefile (.shp)* format, we will load it like so.

```{r}
# Load Data
years <- list('04','07','11','14')
pneuData <- paste0(dir, '/SHPFinal/Neumonia', years, '.shp')
```

And then we will read it using the `sp` package.

```{r, results=FALSE}
# read Data
pneuShp <- lapply(pneuData, st_read)
pneuNames <- as.character(paste0('pneu',years))
names(pneuShp) <- pneuNames
```

# Exploratory Analysis

## Spatial weights matrices

One of the main aspects of any spatial autorregresive model is the contiguity matrix, also known as the spatial weights matrix ($\mathbf{W}$). This matrix encodes the spatial dependence and influence of one region with its neighbors. There are many ways to define $\mathbf{W}$. Usually, an expert can propose a potential spatial weights matrix based on their knowledge of the phenomenon (study variable). For our research, we will consider most of the parametric approaches as we do not make any assumptions about the underlying spatial structure of the SMR variable. 

```{r}
# Contiguity criteria

# Queen
queenW <- lapply(pneuShp, poly2nb)
# Root
rootW <- lapply(pneuShp, poly2nb, queen = F)

# Distance criteria

# Coords
coords <- lapply(pneuShp, function(x) {st_sfc(st_centroid(x$geometry))})

# Delanuay Triangulation
triangW <- lapply(coords, tri2nb)
# Sphere of Influence
sphereW <- lapply(1:length(pneuNames), function(x) {graph2nb(soi.graph(triangW[[x]], coords[[x]]))})
names(sphereW) <- pneuNames
# Gabriel
gabrielW <- lapply(1:length(pneuNames), function(x) {graph2nb(gabrielneigh(coords[[x]]), sym = T)})
names(gabrielW) <- pneuNames
# Relative Neighbors
relativeW <- lapply(1:length(pneuNames), function(x) {graph2nb(relativeneigh(coords[[x]]), sym = T)})
names(relativeW) <- pneuNames

# K-neighbors critreria

# Polygons IDs
IDs <- row.names(as.data.frame(pneuShp$pneu04))

# K-1 neighbor
kn1W <- lapply(1:length(pneuNames), function(x) {knn2nb(knearneigh(coords[[x]], k = 1), row.names = IDs)})
names(kn1W) <- pneuNames

# K-2 neighbors
kn2W <- lapply(1:length(pneuNames), function(x) {knn2nb(knearneigh(coords[[x]], k = 2), row.names = IDs)})
names(kn2W) <- pneuNames

# K-4 neighbors
kn4W <- lapply(1:length(pneuNames), function(x) {knn2nb(knearneigh(coords[[x]], k = 4), row.names = IDs)})
names(kn4W) <- pneuNames
```

Now we plot the matrices to check the relations among regions.

```{r}
# Plot weights matrices
op=par(mfrow=c(3,3), oma=c(0,0,0,0), mar=c(1,1,1,1))
matrices <- c('queen','root', 'triang', 'sphere', 'gabriel', 'relative', 'kn1', 'kn2', 'kn4')

for (w in matrices){
  plot(st_geometry(pneuShp$pneu04), border = 'gray')
  plot(get(paste0(w, 'W'))$pneu04, coords$pneu04, add=T,  pch=20, cex.main=1.5 , col="gray25")
  title(w)
}
```
Now we will evaluate the best matrix for our data. We will use the [Principal coordinates of neighbour matrices (PCNM)](https://www.sciencedirect.com/science/article/abs/pii/S0304380006000925) as our selection criteria. For further details, please refer to the PCNM method.

First we will create a function to extract the AIC of the PCNM method to compare and extract the best matrix per each year.

```{r}
## Spatial weights matrices PCNM

# Function to get AIC for each matrix
getMatrixAIC <- function(x, w){
  pcnm <- test.W(pneuShp[[x]]$SMR, w[[x]])
  aic <- pcnm$all$AICc
  return(aic)
}
```
 
And then we will get the AIC values.

```{r}
# empty list
pcnm <- list()
# Get AICs
for (w in matrices){
  aic <- lapply(1:length(pneuNames), function(x){getMatrixAIC(x,get(paste0(w, 'W')))})
  pcnm[[w]] <- aic
}

# AICs per year and matrix
pcnm <- as.data.frame(do.call(rbind, pcnm))
names(pcnm) <- pneuNames

```
These are the best matrices per year based on the AIC value.

```{r}
# Best weights matrices per year (based on AIC)
bestMatrices <- lapply(1:length(pneuNames),function(x){rownames(pcnm[which.min(pcnm[[x]]),])})
names(bestMatrices) <- pneuNames # 2004: queen, 2007: kn2, 2011: kn4, 2014; kn2
bestMatrices
```



