---
title: "Penumonia SAR Models"
output:
  html_document:
    keep_md: yes
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    number_sections: true
---

This is a reproducible example of the paper [Pneumonia Sar](). In this research, we perform a spatial autorregresive model analysis to measure the spatial influence of socio-economic, enviromental, behavioural and healthcare factors in pneumonia mortality rates. 

# Getting Started

## Packages

First, we will load the packages needed for processing and analyzing our data.

```{r}
# Omitting warmings (be carefull)
options(warn = -1)

# Packages
packages <- c('lmtest','RColorBrewer','classInt','spdep','TeachingDemos','shapefiles','sp','maptools',
             'scatterplot3d','geoR','spatial','fBasics','car','aplpack','RODBC','ggplot2','spgrass6',
             'adespatial','RANN','ade4','olsrr','rgeos','rgdal','spdep','spgwr','GWmodel','nnet','olsrr',
             'stats','classInt','gridExtra','lmtest','car','MASS','caret','glmnet')
```

For loading and/or installing the packages, you can run the following command. The packages that are installed in your machine will be automatically loaded. Those that are not installed will be installed and loaded to your R session. Remember to provide permits if linux is used as operative system.

```{r, results=FALSE, message=FALSE}
# Permits
#Sys.setenv(R_INSTALL_STAGED = FALSE)

# Load packages
for(p in packages){
  if(!require(p,character.only = TRUE)) install.packages(p, dependencies = TRUE)
  library(p,character.only = TRUE)
}
```

## Working directory

It is easier to work in R using a working directory. That way you ensure your inputs and outputs will be loaded and stored in a global directory.

```{r}
# Set working directory
dir <- '/mnt/d/M.Sc. Gesopatial Tecnologies/GeoMundus/GeoMundus 2019/Neumonia'
setwd(file.path(dir))
```

We will also use some useful functions in the exploratory analysis. Load them all from the *SpatialFunctions* folder. Remember this folder has to be in your current working directory.

```{r}
# Load some useful functions
funDir <- file.path(dir, 'SpatialFunctions')
spFun <- list.files(funDir, pattern = '.r', full.names = T)
invisible(lapply(spFun, source))
```

And we will add an extra function for resting the plot options everytime we plot.

```{r}
# Reset par for sequential plotting
resetPar <- function() { dev.new(); op <- par(no.readonly = TRUE); dev.off(); op}
```

# Data

Our data contains the [standardized mortality ratio (SMR)](https://ibis.health.state.nm.us/resource/SMR_ISR.html#:~:text=Standardized%20Mortality%20Ratio%20(SMR)%20is,the%20same%20age%2Fsex%20groups.) of Pneumonia (our study variable) in 19 districts of Bogotá, Colombia for the years 2004, 2007, 2011 and 2014. The data also contains socio-economic, enviromental and healthcare covariates.

As our data is stored in a *shapefile (.shp)* format, we will load it like so.

```{r}
# Load Data
years <- list('04','07','11','14')
pneuData <- paste0(dir, '/SHPFinal/Neumonia', years, '.shp')
```

And then we will read it using the `sp` package.

```{r, results=FALSE}
# read Data
pneuShp <- lapply(pneuData, st_read)
pneuNames <- as.character(paste0('pneu',years))
names(pneuShp) <- pneuNames
```

# Exploratory Analysis

## Spatial weights matrix

One of the main aspects of any spatial autorregresive model is the contiguity matrix, also known as the spatial weights matrix ($\mathbf{W}$). This matrix encodes the spatial dependence and influence of one region with its neighbors. There are many ways to define $\mathbf{W}$. Usually, an expert proposes a potential spatial weights matrix based on their knowledge of the phenomenon (study variable). For our research, we will consider most of the parametric matrices configurations as we do not make any assumptions about the underlying spatial structure of the SMR variable. 

```{r}
# Contiguity criteria

# Queen
queenW <- lapply(pneuShp, poly2nb)
# Root
rootW <- lapply(pneuShp, poly2nb, queen = F)

# Distance criteria

# Coords
coords <- lapply(pneuShp, function(x) {st_sfc(st_centroid(x$geometry))})

# Delanuay Triangulation
triangW <- lapply(coords, tri2nb)
# Sphere of Influence
sphereW <- lapply(1:length(pneuNames), function(x) {graph2nb(soi.graph(triangW[[x]], coords[[x]]))})
names(sphereW) <- pneuNames
# Gabriel
gabrielW <- lapply(1:length(pneuNames), function(x) {graph2nb(gabrielneigh(coords[[x]]), sym = T)})
names(gabrielW) <- pneuNames
# Relative Neighbors
relativeW <- lapply(1:length(pneuNames), function(x) {graph2nb(relativeneigh(coords[[x]]), sym = T)})
names(relativeW) <- pneuNames

# K-neighbors critreria

# Polygons IDs
IDs <- row.names(as.data.frame(pneuShp$pneu04))

# K-1 neighbor
kn1W <- lapply(1:length(pneuNames), function(x) {knn2nb(knearneigh(coords[[x]], k = 1), row.names = IDs)})
names(kn1W) <- pneuNames

# K-2 neighbors
kn2W <- lapply(1:length(pneuNames), function(x) {knn2nb(knearneigh(coords[[x]], k = 2), row.names = IDs)})
names(kn2W) <- pneuNames

# K-4 neighbors
kn4W <- lapply(1:length(pneuNames), function(x) {knn2nb(knearneigh(coords[[x]], k = 4), row.names = IDs)})
names(kn4W) <- pneuNames
```

Now we plot the matrices to check the relationships among regions.

```{r}
# Plot weights matrices
op=par(mfrow=c(3,3), oma=c(0,0,0,0), mar=c(1,1,1,1))
matrices <- c('queen','root', 'triang', 'sphere', 'gabriel', 'relative', 'kn1', 'kn2', 'kn4')

for (w in matrices){
  plot(st_geometry(pneuShp$pneu04), border = 'gray')
  plot(get(paste0(w, 'W'))$pneu04, coords$pneu04, add=T,  pch=20, cex.main=1.5 , col="gray25")
  title(w)
}
```

### Principal coordinates of neighbour matrices (PCNM)

Now we will evaluate the best matrix for our data. We will use the [Principal coordinates of neighbour matrices (PCNM)](https://www.sciencedirect.com/science/article/abs/pii/S0304380006000925) as our selection criteria. For further details, please refer to the PCNM method.

First we will create a function to extract the AIC of the PCNM method to compare and extract the best matrix per each year.

```{r}
## Spatial weights matrices PCNM

# Function to get AIC for each matrix
getMatrixAIC <- function(x, w){
  pcnm <- test.W(pneuShp[[x]]$SMR, w[[x]])
  aic <- pcnm$all$AICc
  return(aic)
}
```
 
And then we will get the AIC values.

```{r, results= FALSE, warning= FALSE}
# empty list
pcnm <- list()
# Get AICs
for (w in matrices){
  aic <- lapply(1:length(pneuNames), function(x){getMatrixAIC(x,get(paste0(w, 'W')))})
  pcnm[[w]] <- aic
}

```
These are the best matrices per year based on the AIC value.

```{r}
# AICs per year and matrix
pcnm <- as.data.frame(do.call(rbind, pcnm))
names(pcnm) <- pneuNames
pcnm

# Best weights matrices per year (based on AIC)
bestMatrices <- lapply(1:length(pneuNames),function(x){rownames(pcnm[which.min(pcnm[[x]]),])})
names(bestMatrices) <- pneuNames # 2004: queen, 2007: kn2, 2011: kn4, 2014; kn2
bestMatrices
```
### Moran's Index

We will confirm the PCNM finding using the Moran's Index of our study variable and the spatial weights matrices. We expect to find spatial autocorrelation between the SMR and $\mathbf{W}$. We will use the Moran's I statistically significant $p-value$ to assess the matrices. 

```{r}

## Compute Moran's I to asses matrices

#set seed
set.seed(123)

# Moran's I test
getMoranPvalue <- function(x, w){
  moran <- moran.mc(pneuShp[[x]]$SMR, nb2listw(w[[x]]), nsim = 1000, zero.policy = TRUE)
  p <- moran$p.value
  return(p)
}

# empty list
moranp <- list()
# Get p values
for (w in matrices){
  pvalue <- lapply(1:length(pneuNames), function(x){getMoranPvalue(x,get(paste0(w, 'W')))})
  moranp[[w]] <- pvalue
}

# p-values per year and matrix
moranp <- as.data.frame(do.call(rbind, moranp))
names(moranp) <- pneuNames # 2004: queen, 2007: kn2, 2011: kn4, 2014: queen
moranp
```

And now having the matrices selected by both the PCNM method and the Moran's I, we chose our final $\mathbf{W}$'s for our study period. We selected the matrices whose PCNM's *AIC* were the lowest and Moran's I $p-values$ were statistically significant.

```{r}
# Selectec matrices based on both PCNM and Moran's I
spatialW <- c('queenW','kn2W', 'kn4W', 'queenW')
```

Now, let us plot the Moran's I of our selected matrices.

```{r}
# Plot Moran's I
par(resetPar())
op=par(mfrow=c(2,2))
for (i in 1:length(spatialW)){
  moran.plot(pneuShp[[i]]$SMR, nb2listw(get(spatialW[[i]])[[i]]), xlab = 'SMR', ylab = 'Spatially lagged SMR', main = paste0('20', years[[i]], '\n ', spatialW[[i]] ))
}
```

### Higher order matrices

We also want to know if our matrices  have statistically significant spatial lags or greater orders, that is, finding autocorrelation throughout longer distances or neighbors orders. If we find statistically significant orders of the matrices, we have to include them in the spatial autorregresive models to guarantee a proper inclusion of the spatial dependence among the regions.

```{r, warning= FALSE, results= FALSE}
# Plot Moran's I correlograms
par(resetPar())
op=par(mfrow=c(2,2))
for (i in 1:length(spatialW)){
  correlogram <- sp.correlogram(get(spatialW[[i]])[[i]], pneuShp[[i]]$SMR, order=5, method= "I",style="W", zero.policy=T)
  print(correlogram)
  plot(correlogram, main = paste0('20', years[[i]], '\n ', spatialW[[i]] ))
} # Besides the first lag, the third lag is significative for 2004 and 2011

```
We see that in 2004 and 2011 the 3rd order matrices are statistically significant. We will add the lagged variable as covariates to include them in the spatial autorregresive (SAR) models.

```{r}
# Define spatial lag variables
lagQueenW3 <- nblag(queenW$pneu04, 3)[[3]]
lagKn4W3 <- nblag(kn4W$pneu11, 3)[[3]]

# Add spatial lags to original data
pneuShp$pneu04$lag3SMR <- lag.listw(nb2listw(lagQueenW3), pneuShp$pneu04$SMR)
pneuShp$pneu11$lag3SMR <- lag.listw(nb2listw(lagKn4W3), pneuShp$pneu11$SMR)
```

## Global Analysis

As an essential condition of the SAR models, the independent variable (SMR), dependent variables (covariates) or distrubances (error term) must exhibit spatial autocorrelation. This spatial autocorrelation can be present across study area (Global) or in specific regions (Local). The former can be found using both the Moran's I and Getis Ord statistics. 

### Global Moran's I

We will assess the presence of global spatial autocorrelation in our independent variable using the global Moran's I.

```{r}
# Moran's I
moranI <- lapply(1:length(pneuNames), function(x) {moran.test(pneuShp[[x]]$SMR, nb2listw(get(spatialW[[x]])[[x]]))})
names(moranI) <- pneuNames
moranI
```
As we can see, all years except 2007 exhibit global spatial autocorrelation using a significance level of $p-value$ < 0.05.

### Goblal Getis Ord

We will corroborate this results using the global statistic Getis Ord statistic.


```{r}
# Getid's Ord
getisO <- lapply(1:length(pneuNames), function(x) {globalG.test(pneuShp[[x]]$SMR, nb2listw(get(spatialW[[x]])[[x]], style = 'B'))})
names(getisO) <- pneuNames
getisO
```

Every year, except 2014, presents positive and statistically significant spatial clustering, that is, there are high values concentration in our study area. The Moran's I informed us of global spatial autocorrelation in 2014. However, it does not determine the existence of clusters of high/low values.

### Bivariate Moran's I

To find spatial association between the independent variable and the covariates, we compute the bivariate Moran's Index. Covariates with statistically significant degree of spatial correlation with the SMR are initial potential candidates for the spatial autorregresive models. We will calculate the Moran's I $p-value$ and plot to scrutinized the covariates that spatially correlate with the SMR. 

The analysis is perform for the entier study period. However, for the sake of brevity, we will display only the results for 2014.

```{r, results = FALSE, warnings = FALSE}
# Change column names (data integrity)
for (y in 1:length(pneuNames)){
  for(col in 77:91){
    colnames(pneuShp[[y]])[col] <-  sub(paste0(years[[y]] ,'.*'), "", colnames(pneuShp[[y]])[col])
  }
}

# Bivariate Moran's I (p-value and plot)
par(resetPar())
op=par(mfrow=c(4,4), mar=c(4,4,1,1),oma=c(1,1,1,1))
for (col in c(78:88, 89:90)){
  mi <- moranbi.test(pneuShp$pneu14$SMR, pneuShp$pneu14[[col]], nb2listw(get(spatialW[[4]])[[4]]), N= 999)
  moranbi.plot(pneuShp$pneu14$SMR, pneuShp$pneu14[[col]], nb2listw(get(spatialW[[4]])[[4]]), N= 999,graph=T, quiet = T, main = paste0('I = ', round(mi$Observed,3), ', p-value = ', mi$p.value), xlab = 'SMR', ylab = colnames(pneuShp$pneu14[col])[1], cex.main=0.9)
} # For 2014 (For other years the variables need to be adressed for each dataset, that is, pneu'year'.  e.g., pneu07)
```





